resevol: an R package for spatially explicit models of pesticide resistance

### 2023-11-30 notebook/ms/ms.Rmd
Some edits to the manuscript prior to acceptance from PLoS Computational Biology

---
title: 'resevol: an R package for spatially explicit models of pesticide resistance
  given evolving pest genomes'
author: |
  | A. Bradley Duthie$^{1,4}$, Rosie Mangan$^{1}$, C. Rose McKeon$^{1}$,
  | Matthew C. Tinsley$^{1}$, and Luc F. Bussi&egrave;re$^{2,3}$
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2]Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden [3] Gothenburgh Global Biodiversity Centre, Gothenburg, Sweden [4] alexander.duthie@stir.ac.uk, Biological and Environmental Sciences 3A149
  University of Stirling Stirling, FK9 4LA, UK'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    fig_caption: yes
    toc: no
  word_document:
    reference_docx: docx_template.docx
  pdf_document:
    toc: no
  html_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[R]{RESEVOL, AN R PACKAGE}
linestretch: 1
link-citations: yes
linkcolor: blue
bibliography: resevol.bib
biblio-style: apalike
---

@@

```{r, echo = FALSE}
library("resevol");
```

@@

Covarying pest quantitative traits {#mine_gmatrix}
===================================================

@@

Conceptually, the relationship between individual genotypes and traits is defined by a network connecting loci to traits through a set of hidden internal nodes (Figure \@ref(fig:network)). 
Values for loci are randomly drawn from a standard normal distribution, $\mathcal{N}\left(0, 1\right)$. 
Links between loci, internal nodes, and traits, can take any real value and are represented by black arrows in Figure \@ref(fig:network). 
Traits can take any real value, and are calculated as the summed effects of all preceding nodes (i.e., the blue squares immediately to the left of the traits in Figure \@ref(fig:network)).
Mathematically, loci are represented by a row vector of length $L$.
Effects of loci on the first layer of internal nodes (black arrows emanating from loci in Figure \@ref(fig:network)) are represented by an $L \times T$ matrix, and transitions between internal nodes, and between the last set of internal nodes and the final traits, are represented by $T \times T$ matrices.
For the example in Figure \@ref(fig:network),

@@

Values mapping loci to traits become part of an individual's genome, so the genome for the individual represented by Figure \@ref(fig:network) is stored in the model as shown below:

@@

Individuals with different loci can therefore have different covarying traits that are constrained by the network structure encoded in each genome. 
Individuals can be haploid (as in Figure \@ref(fig:network)) or diploid (in which case, allele values are summed at homologous loci).

An evolutionary algorithm is used to find appropriate values that produce covarying traits from loci (see Supporting Information 1 for details). 
Evolutionary algorithms are heuristic tools that can simulate adaptive evolution to find solutions for a broad range of problems [@Hamblin2013; @Duthie2018]. 
In the resevol package, the `mine_gmatrix` function runs an evolutionary algorithm and requires the argument `gmatrix`, which specifies the desired trait covariance matrix.
The function initialises a population of `npsize` separate, and potentially unique, networks (i.e., `npsize` copies of a network like the one shown in Figure \@ref(fig:network)), and this population evolves until some maximum iteration (`max_gen`) or minimum expected network stress (`term_cri`) is met. 
In a single iteration of the algorithm, values mutate and crossover occurs between networks (i.e., some of the `npsize` networks swap values with some probability).
Next, trait covariances produced for each network are estimated by initalising `indivs` individuals with loci sampled from a standard normal distribution.
Network stress is calculated as the logged mean squared deviation between estimated covariances and those in `gmatrix`. 
Tournament selection [@Hamblin2013] is then used to determine the networks for the next iteration of the algorithm.
Networks with the lowest estimated stress have the highest fitness, so these networks are disproportionately represented in the next iteration.
Throughout the evolutionary algorithm, the lowest stress network is saved and returned upon network termination.
The robustness of this network's stress to sets of individuals with different loci values can be tested using the `stress_test` function.

An example run of `mine_gmatrix` with the same number of loci and internal nodes as in Figure \@ref(fig:network) is shown below for traits that have an intended covariance of -0.4:

@@

Simulating landscape-level pesticide resistance {#simulating}
===================================================

@@

Landscape {#landscape}
---------------------------------------------------

@@

Pest ecology {#ecology}
---------------------------------------------------

Individual pests can be modelled to have several reproductive systems and life histories.
Pest reproductive system can be specified using the `repro` argument, which accepts `"asexual"` (haploid), `"sexual"` (monoecious), and `"biparental"` (dioecious).
For `"sexual"` pests, the `selfing` argument specifies if self-fertilisation is (`TRUE`) or is not (`FALSE`) allowed.
At the start of a simulation, pests are initialised in a random location.
Initialised pests are of age zero if `rand_age = FALSE` or a random age from zero to `max_age` if `rand_age = TRUE`.
Following initialisation, a single time step proceeds with landscape change (see above), pest aging and metabolism, feeding, pesticide consumption, movement, reproduction, mortality, and immigration (Figure \@ref(fig:flowchart)).
Feeding, pesticide consumption, movement, and reproduction all depend on pest age.
Pests feed and consume pesticide from ages `min_age_feed` to `max_age_feed`, move from ages  `min_age_move` to `max_age_move`, and reproduce from ages `min_age_reproduce` to `max_age_reproduce` (all inclusive).
Food accumulated is lost during aging if `baseline_metabolism > 0` and pest age is within `min_age_metabolism` and `max_age_metabolism`.
The option to set minimum and maximum ages for events makes it possible to model pests with much different life histories [e.g., @Sudo2018].

@@

Pest evolution {#evolution}
---------------------------------------------------

@@

Evolution of the genetic architecture linking loci to traits can be constrained by disabling mutation in genome values linking loci, internal nodes, and traits (i.e., 'network values' represented by arrows in Figure \@ref(fig:network)).
While mutation at loci (green circles in Figure \@ref(fig:network)) is always possible as long as `mutation_pr > 0`, the number of intermediary layers for which network values can mutate is constrained by `net_mu_layers`. 
If `net_mu_layers = 0`, then no network values can mutate, but higher integer values cause mutation to occur at network value layers from loci to traits (if `net_mu_dir = 1`) or traits to loci (if `net_mu_dir = 0`).
For example, if `net_mu_layers = 2` and `net_mu_dir = 1`, then the network values linking loci to the first internal node, and the first internal node to the second, can mutate (i.e., first two columns of arrows in \@ref(fig:network), but not those linking the second internal node to traits).
This allows pest traits to evolve with varying degrees of constraint on the covariance between traits.
Low `net_mu_layers` values model strong genetic constraints, while high values model high evolvability of trait covariances.

@@

Simulation output {#output}
===================================================

@@

Example of individual-based simulations {#simulation}
===================================================

In resevol v0.3, the `run_farm_sim` function includes 68 arguments, which specify a wide range of possible simulation conditions affecting landscape and pest characteristics.
These arguments are explained in the package documentation, and in Supporting Information 2, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
Here we focus on a simple simulation with asexually reproducing pests that have three loci and two traits (Figure \@ref(fig:network)).
We use the the genome generated in `new_network` from [Section 2](#mine_gmatrix), in which traits 1 and 2 have variances of `r new_network[[6]][1,1]` and `r new_network[[6]][2,2]`, respectively, with a covariance of `r new_network[[6]][1,2]`.

@@

Any arguments to `run_farm_sim` not included above are set to default values.
[Output files](#output) can be used to plot ecological and evolutionary dynamics of pests. 
Figure \@ref(fig:overtime)a shows the pest population change both across and within seasons.
Due to the specific parameter values chosen, clear patterns reflecting pest cohorts emerge.
At the start of a season when the most crop is available, pests eat and abundance increases.
As less crop remains in a season, pest abundance decreases.
Mean food consumed varies over the course of a season caused by the varying frequency of pests in different life history stages (Figure \@ref(fig:overtime)c).
In the first 50 time steps, there is some consumption of pesticides 1 and 2 (Figure \@ref(fig:overtime)b), but both traits rapidly evolve to negative values making both pesticides ineffective (Figure \@ref(fig:overtime)d).
In this case, the trade-off in pesticide consumption is not strong enough to maintain susceptibility to either pesticide in the population.
Figure \@ref(fig:landscape) shows the location of pests on the landscape in the last time of the simulation (left panel), and how different pesticides are currently being applied on the landscape (right panel).
Because the number of crops and pesticides, the number and nature of traits, and the size of pest genomes all vary depending on simulation parameters, no plotting functions are introduced in the resevol package.
Instead, methods and code for producing plots such as those in Figures \@ref(fig:overtime) and \@ref(fig:landscape) are explained in Supporting Information 2.

@@

Conclusions {#conclusions}
===================================================

Insecticide resistance to pesticides is a widespread problem that affects global food security [@Tabashnik2014; @Bradshaw2016; @Lykogianni2021]. 
Many models have been developed to investigate the evolution of resistance under different ecological and evolutionary conditions [e.g., @Curtis1985; @Mani1985; @Ives2002; @Liang2013; @Liang2018; @Haridas2018;  @South2018; @Sudo2018; @Saikai2021].
The resevol R package makes it possible to rapidly develop and simulate myriad individual-based models of resistance evolution.
It also introduces a novel approach to modelling complex pest genetic architecture, using an evolutionary algorithm to generate haploid or diploid loci that map to pest traits with pre-specified covariances (Figure \@ref(fig:network); Supporting Information 1).
Agricultural landscapes and pest life histories are highly customisable, allowing targetted models that can simulate specific real-world case studies (Supporting Information 2).
The breadth of possible models that can be simulated with resevol also makes it a useful tool for developing theory on pest management, and even more generally on the evolution and ecology of individuals with complex traits on a heterogeneous landscape.

Acknowledgements {#acknowledgements}
===================================================

This software was developed as part of the project for Enhancing Diversity to Overcome Resistance Evolution (ENDORSE). 
The ENDORSE project is a joint Newton funded international partnership between the Biotechnology and Biological Sciences Research Council (BBSRC) in the UK and the S&#xE3;o Paulo Research Foundation (FAPESP) in Brazil under BBSRC award reference BB/S018956/1 and FAPESP award reference 2018/21089-3. 
ENDORSE is a partnership among Universidade Estadual Paulista (UNESP), the University of Stirling (UoS), and the Centre for Agricultural and Biosciences International (CABI).

@@

Availability {#availability}
===================================================

@@

>>>>

---
title: 'resevol: an R package for spatially explicit models of pesticide resistance
  given evolving pest genomes'
author: |
  | A. Bradley Duthie$^{1,*}$, Rosie Mangan$^{1}$, C. Rose McKeon$^{1}$,
  | Matthew C. Tinsley$^{1}$, and Luc F. Bussi&egrave;re$^{2,3}$
date: '[1] Biological and Environmental Sciences, University of Stirling, Stirling,
  UK [2]Biological and Environmental Sciences, University of Gothenburg, Gothenburg, Sweden [3] Gothenburgh Global Biodiversity Centre, Gothenburg, Sweden [*] alexander.duthie@stir.ac.uk'
output:
  bookdown::pdf_document2:
    citation_package: natbib
    fig_caption: yes
    toc: no
    keep_tex: true
  word_document:
    reference_docx: docx_template.docx
  pdf_document:
    toc: no
  html_document: default
header-includes:
- \usepackage{fancyhdr}
- \pagestyle{fancy}
- \fancyhead[R]{RESEVOL, AN R PACKAGE}
- \usepackage{lineno}
- \linenumbers
- \modulolinenumbers[2]
linestretch: 2
link-citations: yes
linkcolor: blue
csl: plos-computational-biology.csl
bibliography: resevol.bib
biblio-style: acm
---

@@

```{r, echo = FALSE}
library("resevol");
options(tinytex.clean = FALSE);
```

@@

Design and Implementation
===================================================

Covarying pest quantitative traits {#mine_gmatrix}
---------------------------------------------------

@@

Conceptually, the relationship between individual genotypes and traits is defined by a network connecting loci to traits through a set of hidden internal nodes (Fig \@ref(fig:network)). 
Values for loci are randomly drawn from a standard normal distribution, $\mathcal{N}\left(0, 1\right)$. 
Links between loci, internal nodes, and traits, can take any real value and are represented by black arrows in Fig \@ref(fig:network). 
Traits can take any real value, and are calculated as the summed effects of all preceding nodes (i.e., the blue squares immediately to the left of the traits in Fig \@ref(fig:network)).
Mathematically, loci are represented by a row vector of length $L$.
Effects of loci on the first layer of internal nodes (black arrows emanating from loci in Fig \@ref(fig:network)) are represented by an $L \times T$ matrix, and transitions between internal nodes, and between the last set of internal nodes and the final traits, are represented by $T \times T$ matrices.
For the example in Fig \@ref(fig:network),

@@

Values mapping loci to traits become part of an individual's genome, so the genome for the individual represented by Fig \@ref(fig:network) is stored in the model as shown below:

@@

Individuals with different loci can therefore have different covarying traits that are constrained by the network structure encoded in each genome. 
Individuals can be haploid (as in Fig \@ref(fig:network)) or diploid (in which case, allele values are summed at homologous loci).

An evolutionary algorithm is used to find appropriate values that produce covarying traits from loci (see Supporting Information 1 for details). 
Evolutionary algorithms are heuristic tools that can simulate adaptive evolution to find solutions for a broad range of problems [@Hamblin2013; @Duthie2018]. 
In the resevol package, the `mine_gmatrix` function runs an evolutionary algorithm and requires the argument `gmatrix`, which specifies the desired trait covariance matrix.
The function initialises a population of `npsize` separate, and potentially unique, networks (i.e., `npsize` copies of a network like the one shown in Fig \@ref(fig:network)), and this population evolves until some maximum iteration (`max_gen`) or minimum expected network stress (`term_cri`) is met. 
In a single iteration of the algorithm, values mutate and crossover occurs between networks (i.e., some of the `npsize` networks swap values with some probability).
Next, trait covariances produced for each network are estimated by initalising `indivs` individuals with loci sampled from a standard normal distribution.
Network stress is calculated as the logged mean squared deviation between estimated covariances and those in `gmatrix`. 
Tournament selection [@Hamblin2013] is then used to determine the networks for the next iteration of the algorithm.
Networks with the lowest estimated stress have the highest fitness, so these networks are disproportionately represented in the next iteration.
Throughout the evolutionary algorithm, the lowest stress network is saved and returned upon network termination.
The robustness of this network's stress to sets of individuals with different loci values can be tested using the `stress_test` function.

An example run of `mine_gmatrix` with the same number of loci and internal nodes as in Fig \@ref(fig:network) is shown below for traits that have an intended covariance of -0.4:

@@

Simulating landscape-level pesticide resistance {#simulating}
---------------------------------------------------

@@

### Landscape {#landscape}

@@

### Pest ecology {#ecology}

Individual pests can be modelled to have several reproductive systems and life histories.
Pest reproductive system can be specified using the `repro` argument, which accepts `"asexual"` (haploid), `"sexual"` (monoecious), and `"biparental"` (dioecious).
For `"sexual"` pests, the `selfing` argument specifies if self-fertilisation is (`TRUE`) or is not (`FALSE`) allowed.
At the start of a simulation, pests are initialised in a random location.
Initialised pests are of age zero if `rand_age = FALSE` or a random age from zero to `max_age` if `rand_age = TRUE`.
Following initialisation, a single time step proceeds with landscape change (see above), pest aging and metabolism, feeding, pesticide consumption, movement, reproduction, mortality, and immigration (Fig \@ref(fig:flowchart)).
Feeding, pesticide consumption, movement, and reproduction all depend on pest age.
Pests feed and consume pesticide from ages `min_age_feed` to `max_age_feed`, move from ages  `min_age_move` to `max_age_move`, and reproduce from ages `min_age_reproduce` to `max_age_reproduce` (all inclusive).
Food accumulated is lost during aging if `baseline_metabolism > 0` and pest age is within `min_age_metabolism` and `max_age_metabolism`.
The option to set minimum and maximum ages for events makes it possible to model pests with much different life histories [e.g., @Sudo2018].

@@

### Pest evolution {#evolution}

@@

Evolution of the genetic architecture linking loci to traits can be constrained by disabling mutation in genome values linking loci, internal nodes, and traits (i.e., 'network values' represented by arrows in Fig \@ref(fig:network)).
While mutation at loci (green circles in Fig \@ref(fig:network)) is always possible as long as `mutation_pr > 0`, the number of intermediary layers for which network values can mutate is constrained by `net_mu_layers`. 
If `net_mu_layers = 0`, then no network values can mutate, but higher integer values cause mutation to occur at network value layers from loci to traits (if `net_mu_dir = 1`) or traits to loci (if `net_mu_dir = 0`).
For example, if `net_mu_layers = 2` and `net_mu_dir = 1`, then the network values linking loci to the first internal node, and the first internal node to the second, can mutate (i.e., first two columns of arrows in \@ref(fig:network), but not those linking the second internal node to traits).
This allows pest traits to evolve with varying degrees of constraint on the covariance between traits.
Low `net_mu_layers` values model strong genetic constraints, while high values model high evolvability of trait covariances.

@@

Simulation output {#output}
---------------------------------------------------

@@

Results {#simulation}
===================================================

In resevol v0.3, the `run_farm_sim` function includes 68 arguments, which specify a wide range of possible simulation conditions affecting landscape and pest characteristics.
These arguments are explained in the package documentation, and in Supporting Information 2, which demonstrates an advanced case study with a custom landscape and complex pest genomes and life history.
Here we focus on a simple simulation with asexually reproducing pests that have three loci and two traits (Fig \@ref(fig:network)).
We use the the genome generated in `new_network` from [Section 2](#mine_gmatrix), in which traits 1 and 2 have variances of `r new_network[[6]][1,1]` and `r new_network[[6]][2,2]`, respectively, with a covariance of `r new_network[[6]][1,2]`.

@@

Any arguments to `run_farm_sim` not included above are set to default values.
[Output files](#output) can be used to plot ecological and evolutionary dynamics of pests. 
Fig \@ref(fig:overtime)a shows the pest population change both across and within seasons.
Due to the specific parameter values chosen, clear patterns reflecting pest cohorts emerge.
At the start of a season when the most crop is available, pests eat and abundance increases.
As less crop remains in a season, pest abundance decreases.
Mean food consumed varies over the course of a season caused by the varying frequency of pests in different life history stages (Fig \@ref(fig:overtime)c).
In the first 50 time steps, there is some consumption of pesticides 1 and 2 (Fig \@ref(fig:overtime)b), but both traits rapidly evolve to negative values making both pesticides ineffective (Fig \@ref(fig:overtime)d).
In this case, the trade-off in pesticide consumption is not strong enough to maintain susceptibility to either pesticide in the population.
Fig \@ref(fig:landscape) shows the location of pests on the landscape in the last time of the simulation (left panel), and how different pesticides are currently being applied on the landscape (right panel).
Because the number of crops and pesticides, the number and nature of traits, and the size of pest genomes all vary depending on simulation parameters, no plotting functions are introduced in the resevol package.
Instead, methods and code for producing plots such as those in Fig \@ref(fig:overtime) and \@ref(fig:landscape) are explained in Supporting Information 2.

@@

Availability and Future Directions {#conclusions}
===================================================

Insecticide resistance to pesticides is a widespread problem that affects global food security [@Tabashnik2014; @Bradshaw2016; @Lykogianni2021]. 
Many models have been developed to investigate the evolution of resistance under different ecological and evolutionary conditions [e.g., @Curtis1985; @Mani1985; @Ives2002; @Liang2013; @Liang2018; @Haridas2018;  @South2018; @Sudo2018; @Saikai2021].
The resevol R package makes it possible to rapidly develop and simulate myriad individual-based models of resistance evolution.
It also introduces a novel approach to modelling complex pest genetic architecture, using an evolutionary algorithm to generate haploid or diploid loci that map to pest traits with pre-specified covariances (Fig \@ref(fig:network); Supporting Information 1).
Agricultural landscapes and pest life histories are highly customisable, allowing targetted models that can simulate specific real-world case studies (Supporting Information 2).
The breadth of possible models that can be simulated with resevol also makes it a useful tool for developing theory on pest management, and even more generally on the evolution and ecology of individuals with complex traits on a heterogeneous landscape.

@@

Financial Disclosure statement {#financial}
===================================================

This software was developed as part of the project for Enhancing Diversity to Overcome Resistance Evolution (ENDORSE) led by Luc Bussi&egrave;re, Ricardo Polanczyk, and Matthew Tinsley (Co-Investigators: Nils Bunnefeld, Yelitza Colmenarez, Nat&aacute;lia Corniani, Renata de Lima, Brad Duthie, Steve Edgington, Leonardo Fraceto, Belinda Luke, and Rosie Mangan). 
The ENDORSE project is a joint Newton funded international partnership between the Biotechnology and Biological Sciences Research Council (BBSRC) in the UK and the S&atilde;o Paulo Research Foundation (FAPESP) in Brazil under BBSRC award reference BB/S018956/1 and FAPESP award reference 2018/21089-3. 
ENDORSE is a partnership among Universidade Estadual Paulista (UNESP), the University of Stirling (UoS), and the Centre for Agricultural and Biosciences International (CABI). 
The funders had no role in study design, data collection and analysis, decision to publish, or preparation of the manuscript.

Competing interests {#competing}
===================================================

The authors have declared that no competing interests exist.

@@

Supporting information {#SI}
===================================================

S1 Text. The evolutionary algorithm of the resevol R package, including key data structures used, a general overview of the evolutionary algorithm, and details concerning haploid and diploid individuals. (PDF)

S2 Text. Advanced techniques for using the resevol R package, including initialising pest genomes and running simulations. (PDF)
